MSSM-Penn-Pitt - ACC and DLPFC - DREAM - SALMON - GRCh38 - Model Selection

### 2020-04-01 cmc-rnaseq-de/MSSM-Penn-Pitt_ACC_DLPFC_SALMON_GRCh38.Rmd
modify inputs for salmon transcripts

rownames(counts) <- counts$Geneid
counts = counts[,indToRetain]

@@

```{r gene.param}
## Get GC content from biomart
backgroundGenes = data.frame(gene_id = rownames(counts)) %>%
  dplyr::mutate(id = gene_id) %>%
  tidyr::separate(id, c('ensembl_gene_id','position'), sep = '\\.')

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>% 
  dplyr::select(gene_id, percentage_gene_gc_content, chromosome_name) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$gene_id
```

### Sex Check
```{r Reported_Gender_plots}
counts$gene_id = rownames(counts) 
REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
  left_join(counts) %>%
  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
  tidyr::gather(key = item, value = value, -c(gene_id, chromosome_name)) %>%
  dplyr::mutate(value = log(value)) %>%
  dplyr::rename(`counts(log)`= value) %>% 
  dplyr::rename(SampleID = item) %>%
  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
  dplyr::rename(`Reported Gender` = Reported_Gender)

##XIST and UTY expression 
#ENSG00000229807.11 and ENSG00000183878.15 
FILT <- REPORTED.GENDER.COUNTS %>% 
  filter(gene_id == "ENSG00000229807.11" | gene_id == "ENSG00000183878.15") %>% 
  dplyr::select(-one_of("chromosome_name")) %>% 
  tidyr::spread(key = gene_id, value = `counts(log)`) %>% 
  mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
  mutate(UTY = as.numeric(`ENSG00000183878.15`))

@@

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Remove genes with percentage GC content
```{r cpmnormalisation}
counts$gene_id = NULL

@@

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC.CONT$gene_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)])

#Subset union of all genes and convert subset to logcpm values 
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(counts[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
  dplyr::rename(gene_id = genes) %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  group_by(gene_biotype) %>%
  dplyr::summarise(fraction  = n()) %>%
  filter(fraction > 100) %>%
  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
```
`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.

@@

# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_gene_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

@@

                                weights = voomWeights[i,]
                                BIC = NA
                                tryCatch({
                                  mdl = lmer(formula = formula, 
                                             data = data, 
                                             weights = weights)
                                  BIC = BIC(mdl)
                                }, error = function(e){
                                  BIC = NA
                                })
                              return(data.frame(ensembl_gene_id = rownames(cnt)[i],
                                                BIC = BIC))
                            }
  return(modelBIC)
  }
# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, cnt, baseModelBIC){

@@

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -ensembl_gene_id) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = better - worse)

@@

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(ensembl_gene_id, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

@@

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_gene_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(COUNT_OBJ, annotations = all.annotations)
```

>>>>

counts = counts[,indToRetain]

@@

```{r gene.param}
## Get GC content from biomart
backgroundGenes = data.frame(transcript_id = rownames(counts)) %>%
  dplyr::mutate(id = transcript_id) %>%
  tidyr::separate(id, c('ensembl_transcript_id','position'), sep = '\\.')

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_transcript_id", "hgnc_trans_name", "percentage_gene_gc_content", "gene_biotype", "chromosome_name"),
                       filters = "ensembl_transcript_id", values = backgroundGenes$ensembl_transcript_id,
                       mart = mart)

GENE.GC.CONT = Ensemble2HGNC %>%
  dplyr::left_join(backgroundGenes) %>% 
  dplyr::select(transcript_id, percentage_gene_gc_content, chromosome_name) %>%
  unique
rownames(GENE.GC.CONT) = GENE.GC.CONT$transcript_id
```

### Sex Check
```{r Reported_Gender_plots}
counts$transcript_id = rownames(counts) 
REPORTED.GENDER.COUNTS = GENE.GC.CONT %>% 
  left_join(counts) %>%
  dplyr::select(-one_of("percentage_gene_gc_content")) %>%
  filter(chromosome_name == "X" |chromosome_name == "Y") %>% 
  tidyr::gather(key = item, value = value, -c(transcript_id, chromosome_name)) %>%
  dplyr::mutate(value = log(value)) %>%
  dplyr::rename(`counts(log)`= value) %>% 
  dplyr::rename(SampleID = item) %>%
  left_join(METADATA[,c("SampleID", "Reported_Gender")]) %>% 
  dplyr::rename(`Reported Gender` = Reported_Gender)

##XIST and UTY expression 
#ENSG00000229807.11 and ENSG00000183878.15 
FILT <- REPORTED.GENDER.COUNTS %>% 
  filter(transcript_id == "ENSG00000229807.11" | transcript_id == "ENSG00000183878.15") %>% 
  dplyr::select(-one_of("chromosome_name")) %>% 
  tidyr::spread(key = transcript_id, value = `counts(log)`) %>% 
  mutate(XIST = as.numeric(`ENSG00000229807.11`)) %>% 
  mutate(UTY = as.numeric(`ENSG00000183878.15`))

@@

# Gender
# vcd::mosaic(~ Gender + Dx.Tissue, data = COVARIATES)
```
### Filter genes
Remove genes that have less than 1 cpm counts in at least 50% of samples per Dx.Tissue. Remove genes with percentage GC content
```{r cpmnormalisation}
counts$transcript_id = NULL

@@

genesToAnalyze = unlist(genesToAnalyze) %>% 
  unique() %>%
  intersect(GENE.GC.CONT$transcript_id[!is.na(GENE.GC.CONT$percentage_gene_gc_content)])

#Subset union of all genes and convert subset to logcpm values 
PROCESSED_COUNTS = getGeneFilteredGeneExprMatrix(counts[genesToAnalyze, ], MIN_GENE_CPM=0, MIN_SAMPLE_PERCENT_WITH_MIN_GENE_CPM=0)
pct.pc = PROCESSED_COUNTS$filteredExprMatrix$genes %>%
  dplyr::rename(transcript_id = genes) %>%
  left_join(backgroundGenes) %>%
  left_join(Ensemble2HGNC) %>%
  group_by(gene_biotype) %>%
  dplyr::summarise(fraction  = n()) %>%
  filter(fraction > 100) %>%
  dplyr::mutate(fraction = fraction/length(PROCESSED_COUNTS$filteredExprMatrix$genes[,1]))
```
`r dim(PROCESSED_COUNTS$filteredExprMatrix$genes)[1]` genes are used for the analysis.

@@

# Plot abberent distribution of logcpm counts
tmp1 = voom(PROCESSED_COUNTS$filteredExprMatrix$counts)$E %>%
  rownameToFirstColumn('ensembl_transcript_id') %>%
  tidyr::gather(SampleID, logCPM, -ensembl_transcript_id) %>%
  left_join(COVARIATES %>%
              rownameToFirstColumn('SampleID') %>%
              tidyr::separate(Dx.Tissue, c('Dx', 'Tissue'), sep = '_'))

@@

                                weights = voomWeights[i,]
                                BIC = NA
                                tryCatch({
                                  mdl = lmer(formula = formula, 
                                             data = data, 
                                             weights = weights)
                                  BIC = BIC(mdl)
                                }, error = function(e){
                                  BIC = NA
                                })
                              return(data.frame(ensembl_transcript_id = rownames(cnt)[i],
                                                BIC = BIC))
                            }
  return(modelBIC)
  }
# Function to select covariates in a step-wise regression fashion
selectCovarsToAdjust <- function(covarsToSelect, adjustCovars, covars, cnt, baseModelBIC){

@@

    # Find number of genes that are changed based on BIC
    nChange = allModelBIC %>%
      tidyr::gather(ModelName, Value, -ensembl_transcript_id) %>%
      dplyr::left_join(baseModelBIC) %>%
      dplyr::mutate(diff = baseModel - Value) %>%
      dplyr::group_by(ModelName) %>%
      dplyr::summarise(better = sum(diff >= 5), 
                       worse = sum(diff <= -5)) %>%
      dplyr::mutate(nchange = better - worse)

@@

    if (dim(tmp)[1] == 0){
      covarsToSelect = c()
    } else {
      baseModelBIC = allModelBIC %>%
        dplyr::select(ensembl_transcript_id, one_of(tmp$ModelName))
      colnames(baseModelBIC)[2] = 'baseModel'
      varsSelected = c(varsSelected, tmp$ModelName)
      covarsToSelect = setdiff(covarsToSelect, tmp$ModelName)
      print(paste('Variable chosen to be added:',tmp$ModelName))
    }
  }

@@

# Store filtered counts
PROCESSED_COUNTS$filteredExprMatrix$counts %>%
  rownameToFirstColumn('ensembl_transcript_id') %>%
  write.table(file = 'CMC_ACC_DLPFC_Counts.tsv', sep = '\t', row.names=F, quote=F)
COUNT_OBJ = File('CMC_ACC_DLPFC_Counts.tsv', name = 'Counts (filtered raw)', parentId = CODE$properties$id)
COUNT_OBJ = synStore(COUNT_OBJ, used = ALL_USED_IDs, activityName = activityName, 
                      executed = thisFile, activityDescription = activityDescription)
synSetAnnotations(COUNT_OBJ, annotations = all.annotations)
```
